### 2019-11-18 ic-journal/response-letter2.tex

\newcommand{\Mark}{{\underline{\tt R:} }}

\begin{center}
\textbf{
{\small Changes for Submission 1 to SI:EXPRESS/SOS'18 (2nd Round)}\\[5mm]
{\large Unique Solutions of Contractions, CCS, and their HOL
  Formalisation
}}
\end{center}

\vskip 10pt
\noindent

(TODO)
% We would like to thank the reviewers for many useful
% comments and suggestions. We believe we have taken all of them into consideration, as best
% as we could.

% The main addition, here again following the recommendations of the reviewers 
% has been the extension of the our central theorems 
% to the multivariable case (ie., moving from ``unique solution theorems'' for a single
% equations, or contractions,  to multi-equations  or multi-contractions). 
%  This has been a significant effort (also because a few alternative ways to make the
%  extension had to be examined). At the end, this work 
% consists of another 4,000 lines of proof scripts and a whole new
% Section 5 in the paper.

In the following, we discuss the changes to the paper and provide a
point-by-point reply to the issues that have been raised.

Our answers are marked  with a `` \Mark''  at the beginning.

\ \\

Best regards,
\begin{flushright}
  Chun Tian and Davide Sangiorgi \\
  (TODO) December 22, 2019
\end{flushright}

\vspace{1cm}

\section*{Responses to Review 1}

\subsection*{Meta comments}

\begin{enumerate}

\item \begin{quote}
    The multivariable case has been added. This is great: it makes the result stronger, and adds a real improvement over the conference version. 
  \end{quote}
  \Mark
  Thanks.

\item \begin{quote}
Overall the authors did address the minor comments. But the paper
still contains quite a few grammatical and other minor mistakes, also
in the newly added text. Below are a couple that I found just browsing
again through the paper; but this is certainly inexhaustive and the
authors should do some more careful proofreading.
\end{quote}
\Mark
(TODO) We have done a more careful proofreading and fixed all
grammatically wrong sentences in the paper.

\item \begin{quote}
    In fact, one of the corrections appears to have introduced a new
    bug: Theorem 4.7 in the new version is wrong; I guess Def. 4.6
    should be about weak transitions. In the original submission,
    there was verbatim copied proof code, which was however
    correct. But as it stands now, it suffers from the famous problem
    with weak bisimulation up-to found by the second author (one can
    simply use the same counterexample $\{(\tau. a , 0)\}$).
  \end{quote}
  \Mark
  This is an embarassing typo. The two $\arr{\mu}$ in
  Definition 4.6 should be $\Arr{\mu}$ instead. Our actual formalization is
  correct. Below is the fixed part of Definition 4.6:
\begin{enumerate}
\item Whenever $P \Arr{\mu} P'$ then, for some $Q'$, $Q \Arcap{\mu} Q'$ and $P' \wb \S \wb Q'$,
\item Whenever $Q \Arr{\mu} Q'$ then, for some $P'$, $P \Arcap{\mu} P'$ and $P' \wb \S \wb Q'$.
\end{enumerate}

\end{enumerate}

\subsection*{Detailed comments}

### 2019-11-18 ic-journal/response-letter2.tex

\item \begin{quote}
    In fact, one of the corrections appears to have introduced a new
    bug: Theorem 4.7 in the new version is wrong; I guess Def. 4.6
    should be about weak transitions. In the original submission,
    there was verbatim copied proof code, which was however
    correct. But as it stands now, it suffers from the famous problem
    with weak bisimulation up-to found by the second author (one can
    simply use the same counterexample $\{(\tau. a , 0)\}$).
  \end{quote}
  \Mark
  This is an embarassing typo. The two $\arr{\mu}$ in
  Definition 4.6 should be $\Arr{\mu}$ instead. Our actual formalization is
  correct. Below is the fixed part of Definition 4.6:
\begin{enumerate}
\item Whenever $P \Arr{\mu} P'$ then, for some $Q'$, $Q \Arcap{\mu} Q'$ and $P' \wb \S \wb Q'$,
\item Whenever $Q \Arr{\mu} Q'$ then, for some $P'$, $P \Arcap{\mu} P'$ and $P' \wb \S \wb Q'$.
\end{enumerate}

\end{enumerate}

\subsection*{Detailed comments}

\begin{enumerate}
  \item \begin{quote}
25 "essentially formalised" => "essentially formalising" ? something is grammatically off with this sentence
\end{quote}
  \Mark

  \item \begin{quote}
47 ", then consider" => ". Then we consider" or something like that
\end{quote}
  \Mark

  \item \begin{quote}
311 "break the logic consistencies" break the logical consistency?
\end{quote}
  \Mark

  \item \begin{quote}
337 "rec" is blue
\end{quote}
  \Mark Fixed. The ``rec'' was wrongly identified as the keyword of ML
  languages in the \LaTeX{} verbatim package that we used.

  \item \begin{quote}
323 and around: why is "a" in italics and $\bar{b}$ not?
\end{quote}
  \Mark

  \item \begin{quote}
sec 4.2: I'm now totally lost on the meaning of $(\alpha, \beta) CCS$. First $\beta$ is a label, then an action, and then there is some mystifying text between brackets (l328) about how this is not explained. *Please* just explain clearly what $\alpha$ and $\beta$ mean in $(\alpha, \beta) CCS$. 
\end{quote}
  \Mark

  \item \begin{quote}
370 spurious \texttt{CCS_Subst} on line 370
\end{quote}
  \Mark

  \item \begin{quote}
530 "for the contraction" the contraction what? the game? functional? or remove "the"
\end{quote}
  \Mark

  \item \begin{quote}
555 "an “up-to” techniques." => "“up-to” techniques."
\end{quote}
  \Mark

  \item \begin{quote}
566 "the resulting ``weak bisimulation up``" correct; also the `` at the right should be ''
\end{quote}
  \Mark

  \item \begin{quote}
597 "not powerful *enough* to prove" 
\end{quote}
  \Mark

  \item \begin{quote}
600 perhaps point out that this is (almost) 
\end{quote}
  \Mark

  \item \begin{quote}
608 "This theorem is of special interests to us, because within our framework it seems impossible to prove it without any limitation" interests => interest; also, I do not really understand this sentence (which limitation?)
\end{quote}
  \Mark

  \item \begin{quote}
965 "Recall *from* the beginning of Section 3 *that*"
\end{quote}
  \Mark

  \item \begin{quote}
985 Is there anything fundamentally interesting going on in the proof? Otherwise perhaps remove it or replace by a high-level description?
\end{quote}
  \Mark

  \item \begin{quote}
1066 "the proof cannot complete" ?
\end{quote}
  \Mark

  \item \begin{quote}
1075 "substitutes each possibly occurrences" "substitues each possible occurence"
\end{quote}
  \Mark

  \item \begin{quote}
page 30: some the text in these lemma's is essentially (informal) explanation, with the HOL code giving the formal statements. Perhaps move the text out of the lemma environment? This will also make it look better, as it's now a full page of lemma's. 
\end{quote}
  \Mark

  \item \begin{quote}
1149: "Recall *that* in"
\end{quote}
  \Mark

  \item \begin{quote}
1160: So the idea is that, whenever we replace a variable by a hole and view the other variables as constants, this is a single variable context - right? Perhaps this (expressed in line 960) can be emphasised again around here (but not entirely sure). 
\end{quote}
  \Mark

  \item \begin{quote}
1165: "used *to* assert"
\end{quote}
  \Mark

  \item \begin{quote}
1169 add colon after e.g.
\end{quote}
  \Mark

  \item \begin{quote}
1171 add "then" after the comma 
\end{quote}
  \Mark

  \item \begin{quote}
1175 "The similar properties" => "Similar properties"
\end{quote}
  \Mark

  \item \begin{quote}
1182 "Not every expression fit with CCS syntax " ?
\end{quote}
  \Mark

  \item \begin{quote}
1210 "optional restriction" ?
\end{quote}
  \Mark

  \item \begin{quote}
1219 "that, the set" remove comma; also "must be disjoint" => "is disjoint"
\end{quote}
  \Mark

  \item \begin{quote}
1241 "while Milner did not pointed out explicitly," grammar
\end{quote}
  \Mark

  \item \begin{quote}
1293 " for people who can read them" sounds a bit strange; also the last sentence of the paragraph suggest most people could
\end{quote}
  \Mark

  \item \begin{quote}
1293-1294 "For some other textbook proofs (..) formalisation work." perhaps give a pointer to an instance where this was the case?
\end{quote}
  \Mark

### 2019-11-18 ic-journal/formal.tex
Minor fixes, closing the paper
Final fixes; disabled highlights
Clean up some old (confirmed) highlights

We highlight here a formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, \hl{with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems} mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
%  The main purpose is to convince the readers that, there's no flaw
%  in the informal proofs. 
% \hl{The univariate versions of these theorems, apart from minor fixes and
% extensions in this paper, was already described in}
% \cite{Tian:2017wrba}. The multivariate proofs and the related results
% are new, and will be described in Section~\ref{sec:multivariate}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to LCF
\cite{gordon1979edinburgh,milner1972logic} by Robin Milner and others
since \hl{1972. It} is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in HOL family (e.g. Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed the HOL has \hl{considerably} simpler logical
foundations than most other theorem provers. %, e.g. Coq. 
\hl{As a consequence, theories and proofs verified in HOL are easier to understand
 to people who are not familar with more advanced
dependent type theories.}

HOL4 is implemented in Standard ML, the same programming language who
plays three different roles:
\begin{enumerate}
\item the underlying implementation language for the core HOL engine;
\item \hl{the language in which proof tactics are implemented;}
\item \hl{the interface  language of the HOL proof scripts and
  interactive shell.}
\end{enumerate}
Moreover, using the same language HOL4 users can write complex automatic
verification tools by calling HOL's theorem proving
facilities. The formal proofs of theorems in CCS theory
are mostly done by an \emph{interactive process} closely following
their informal proofs, with minimal automatic proof searching.

@@

The deductive system of the HOL is specified by eight primitive
derivation rules: (c.f.~\cite{hollogic} for more details.)
\begin{enumerate}
\item Assumption introduction (\texttt{ASSUME});
\item Reflexivity (\texttt{REFL});
\item $\beta$-conversion (\texttt{BETA\_CONV});
\item Substitution (\texttt{SUBST});
\item Abstraction (\texttt{ABS});
\item Type instantiation (\texttt{INST\_TYPE});
\item Discharging an assumption (\texttt{DISCH});
\item Modus Ponens (\texttt{MP}).
\end{enumerate}
All proofs are eventually reduced to applications of the above rules,
which also give the semantics of two foundamental
logical connectives, the equality ($=$) and implication
($\Rightarrow$). The rest logical connectives and first-order
quantifiers, including the logical true (\HOLinline{\HOLConst{T}}) and false (\HOLinline{\HOLConst{F}}), are
further defined as $\lambda$-functions: (one rarely needs to care
things at this level, however.)
\begin{equation*}
\begin{array}{l}
\turn \T       =  ((\lquant{x_{\ty{bool}}}x) =
               (\lquant{x_{\ty{bool}}}x))    \\
\turn \forall  =  \lquant{P_{\alpha\fun\ty{bool}}} P =
                    (\lquant{x}\T ) \\
\turn \exists  =  \lquant{P_{\alpha\fun\ty{bool}}} P({\hilbert}\, P) \\
\turn \F  =  \uquant{b_{\ty{bool}}} b  \\
\turn \neg    =  \lquant{b} b \imp \F \\
\turn {\wedge}  =  \lquant{b_1\ b_2}\uquant{b} (b_1\imp (b_2 \imp b)) \imp b \\
\turn {\vee}  =  \lquant{b_1\ b_2}\uquant{b} (b_1 \imp b)\imp ((b_2 \imp b) \imp b) \\
\turn \OneOne  =  \lquant{f_{\alpha \fun\beta}}\uquant{x_1\ x_2}
                    (f\ x_1 = f\ x_2)  \imp (x_1 = x_2) \\
\turn \Onto  =  \lquant{f_{\alpha\fun\beta}}
                  \uquant{y}\equant{x} y = f\ x \\
\turn \TyDef  =  \lambda P_{\alpha\fun\ty{bool}}\
                  rep_{\beta\fun\alpha}.\;
                  \OneOne\ rep \ \wedge{}\  (\uquant{x}P x = (\equant{y} x = rep\ y))
\end{array}
\end{equation*}
It deserves to mention that, the last logical constant above,
$\TyDef$, can be used to define new HOL types as bijections of
subsets of existing types. (c.f.~\cite{Melham:1989dk} for more details.)
HOL's \texttt{Datatype} package~\cite{Melham:1991, holdesc} automates
this tedious process, and can be used to define needed types in CCS.

@@

Notice that, usually the above four axioms are the only axioms allowed
in conventional formalisation projects in HOL: adding new axioms may
either break the logic consistencies or make the whole work less
convincible. The theorem prover (in general) guarantees that,
without adding new axioms, if both the definitions and the statements
of theorems are correct, then any proof which successfully completes must
be correct.

>>>>

We highlight here a formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to LCF
\cite{gordon1979edinburgh,milner1972logic} by Robin Milner and others
since 1972. It is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in HOL family (e.g. Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed the HOL has considerably simpler logical
foundations than most other theorem provers. %, e.g. Coq. 
As a consequence, theories and proofs verified in HOL are easier to understand
 to people who are not familar with more advanced
dependent type theories.

HOL4 is implemented in Standard ML, the same programming language who
plays three different roles:
\begin{enumerate}
\item \hl{The} underlying implementation language for the core HOL engine;
\item \hl{The} language in which proof tactics are implemented;
\item \hl{The} interface language of the HOL proof scripts and interactive shell.
\end{enumerate}
Moreover, using the same language HOL4 users can write complex automatic
verification tools by calling HOL's theorem proving
facilities. The formal proofs of theorems in CCS theory
are mostly done by an \emph{interactive process} closely following
their informal proofs, with minimal automatic proof searching.

@@

The deductive system of the HOL is specified by eight primitive
derivative rules: (c.f.~\cite{hollogic} for more details.)
\begin{enumerate}
\item Assumption introduction (\texttt{ASSUME});
\item Reflexivity (\texttt{REFL});
\item $\beta$-conversion (\texttt{BETA\_CONV});
\item Substitution (\texttt{SUBST});
\item Abstraction (\texttt{ABS});
\item Type instantiation (\texttt{INST\_TYPE});
\item Discharging an assumption (\texttt{DISCH});
\item Modus Ponens (\texttt{MP}).
\end{enumerate}
All proofs are eventually reduced to applications of the above rules,
which also give the semantics of two foundamental
logical connectives, the equality ($=$) and implication
($\Rightarrow$). The rest logical connectives and first-order
quantifiers, including the logical true (\HOLinline{\HOLConst{T}}) and false (\HOLinline{\HOLConst{F}}), are
further defined as $\lambda$-functions: (one rarely needs to care
things at this level, however.)
\begin{equation*}
\begin{array}{l}
\turn \T       =  ((\lquant{x_{\ty{bool}}}x) =
               (\lquant{x_{\ty{bool}}}x))    \\
\turn \forall  =  \lquant{P_{\alpha\fun\ty{bool}}} P =
                    (\lquant{x}\T ) \\
\turn \exists  =  \lquant{P_{\alpha\fun\ty{bool}}} P({\hilbert}\, P) \\
\turn \F  =  \uquant{b_{\ty{bool}}} b  \\
\turn \neg    =  \lquant{b} b \imp \F \\
\turn {\wedge}  =  \lquant{b_1\ b_2}\uquant{b} (b_1\imp (b_2 \imp b)) \imp b \\
\turn {\vee}  =  \lquant{b_1\ b_2}\uquant{b} (b_1 \imp b)\imp ((b_2 \imp b) \imp b) \\
\turn \OneOne  =  \lquant{f_{\alpha \fun\beta}}\uquant{x_1\ x_2}
                    (f\ x_1 = f\ x_2)  \imp (x_1 = x_2) \\
\turn \Onto  =  \lquant{f_{\alpha\fun\beta}}
                  \uquant{y}\equant{x} y = f\ x \\
\turn \TyDef  =  \lambda P_{\alpha\fun\ty{bool}}\
                  rep_{\beta\fun\alpha}.\;
                  \OneOne\ rep \ \wedge{}\  (\uquant{x}P x = (\equant{y} x = rep\ y))
\end{array}
\end{equation*}
It deserves to mention that, the last logical constant above,
$\TyDef$, can be used to define new HOL types as bijections of
subsets of existing types. (c.f.~\cite{Melham:1989dk} for more details.)
HOL's \texttt{Datatype} package~\cite{Melham:1991, holdesc} automates
this tedious process, and can be used to define needed types in CCS.

@@

Notice that, usually the above four axioms are the only axioms allowed
in conventional formalisation projects in HOL: adding new axioms may
either break the logic consistencies or make the whole work less
convincible. The theorem prover (in general) guarantees that,
without adding new axioms, if both the definitions and the statements
of theorems are correct, then any proof which successfully completes must
be correct. This is also the case here: SOS rules of CCS are
\emph{not} axioms but consequences of the inductive relation\hl{, the
  CCS transition $\rightarrow$}.

### 2019-11-18 ic-journal/intro.tex
Clean up some old (confirmed) highlights

This paper presents a comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
Many results in Milner's CCS book~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
bisimilarities (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma and Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction defining package.

@@

We first consider the case of a single equation (or contraction),
that we call the \univariate case, then consider the \multivariate
  case, where  more than one equation
  (or contraction) with multiple equation variables is considered.
The \univariate theorems use
$\lambda$-functions to present CCS equations,
while the \multivariate versions require a
more careful and delicate treatment of CCS
expressions with multiple variables and substitutions acting on them.

>>>>

This paper presents a comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
Many results in Milner's CCS book~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
bisimilarities (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma and Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially \hl{formalising} van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction defining package.

@@

We first consider the case of a single equation (or contraction)
that we call the \emph{\univariate case}. \hl{Then we} consider
the \emph{\multivariate case}, where more than one equation
  (or contraction) with multiple equation variables is considered.
The \univariate versions \hl{of unique-solution theorems} use
$\lambda$-functions to present CCS equations,
while the \multivariate versions require a
more careful and delicate treatment of CCS
expressions with multiple variables and substitutions acting on them.
